---
title: "Initial EDA: Wildfire School Closures and Absenteeism by Reason"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: false
    fig-width: 8
    fig-height: 5
  pdf:
    fig-width: 7
    fig-height: 5
execute:
  warning: false
  message: false
editor: visual
---

## Setup

This EDA follows three questions:

1.  Are wildfires a major driver of instructional disruption?
2.  Is the closure burden distributed unevenly across counties, districts, and schools?
3.  What do absenteeism-by-reason patterns look like in schools with wildfire closure burden?

The closure dataset uses `year` as the ending year of the academic year. For example, `2019` represents the 2018-19 academic year. The absenteeism analysis therefore uses CDE Absenteeism by Reason files for 2017-18 and 2018-19, matched to closure `year == 2018` and `year == 2019`. The 2019-20 absenteeism file is intentionally excluded because CDE marks it unavailable/not valid due to COVID-era disruption.

`days_quarantine` and `days_safety_other` appear to be subcategories of student safety in the codebook, so they are not included in the main top-level closure-cause totals below to avoid double counting.

```{r setup}
library(tidyverse)
library(janitor)
library(scales)

theme_set(theme_minimal(base_size = 12))

to_number <- function(x) {
  readr::parse_number(as.character(x), na = c("", "NA", "N/A", "NULL", "*"))
}

safe_divide <- function(num, den) {
  if_else(!is.na(den) & den > 0, num / den, NA_real_)
}

clean_cds_code <- function(x) {
  digits <- str_remove_all(coalesce(as.character(x), ""), "\\D")
  if_else(digits == "", NA_character_, str_pad(digits, width = 14, side = "left", pad = "0"))
}

pad_code <- function(x, width) {
  digits <- str_remove_all(coalesce(as.character(x), ""), "\\D")
  if_else(digits == "", NA_character_, str_pad(digits, width = width, side = "left", pad = "0"))
}

year_breaks <- function(x) {
  seq(min(x, na.rm = TRUE), max(x, na.rm = TRUE), by = 2)
}

analysis_years <- c(2018, 2019)
analysis_year_labels <- c("2017-18", "2018-19")

cause_palette <- c(
  "Wildfire" = "#C7432B",
  "Natural disasters/weather" = "#3F7D8C",
  "Infrastructure" = "#D0A03A",
  "Student safety" = "#5E6E46",
  "Other" = "#7D7A78"
)

burden_palette <- c(
  "No wildfire" = "#BFC3C7",
  "Affected: lower 75%" = "#E7A15A",
  "Affected: top 25%" = "#C7432B"
)
```

## Load closure data

```{r load-closure-data}
data_file <- "Annual school closures and standardized test scores in California, 2003-2019.csv"
data_path_candidates <- c(
  file.path("..", "school_fire_data", data_file),
  file.path("school_fire_data", data_file)
)

data_path <- data_path_candidates[file.exists(data_path_candidates)][1]
if (is.na(data_path)) {
  stop("Could not find the closure CSV. Check the relative path from this Quarto file.")
}

df <- readr::read_csv(
  data_path,
  col_types = readr::cols(
    cds_code = readr::col_character(),
    .default = readr::col_guess()
  ),
  show_col_types = FALSE
) |>
  clean_names()

# The provided column listing uses days_wildfires, while the codebook uses
# Days.Wildfire. Standardize to singular so downstream code is consistent.
if ("days_wildfires" %in% names(df) && !"days_wildfire" %in% names(df)) {
  df <- df |>
    rename(days_wildfire = days_wildfires)
}

primary_cause_vars <- c(
  "days_wildfire",
  "days_natural_disasters_weather",
  "days_infrastructure",
  "days_student_safety",
  "days_other"
)

closure_numeric_vars <- c(
  primary_cause_vars,
  "days_quarantine",
  "days_safety_other",
  "days_total",
  "total_enrollment"
)

df <- df |>
  mutate(
    year = as.integer(year),
    closure_year = year,
    academic_year = case_when(
      !is.na(year) ~ str_c(year - 1, "-", str_sub(as.character(year), 3, 4)),
      TRUE ~ NA_character_
    ),
    county = str_squish(str_to_title(county)),
    district_name = str_squish(district_name),
    school_site = str_squish(school_site),
    cds_code_raw = cds_code,
    cds_code = clean_cds_code(cds_code),
    across(any_of(closure_numeric_vars), to_number),
    across(any_of(primary_cause_vars), ~ replace_na(.x, 0)),
    days_total = replace_na(days_total, 0),
    total_enrollment = replace_na(total_enrollment, 0),
    wildfire_student_days = days_wildfire * total_enrollment,
    wildfire_share_of_total = safe_divide(days_wildfire, days_total),
    any_wildfire = days_wildfire > 0
  )

cause_lookup <- tibble(
  cause = primary_cause_vars,
  cause_label = factor(
    c(
      "Wildfire",
      "Natural disasters/weather",
      "Infrastructure",
      "Student safety",
      "Other"
    ),
    levels = c(
      "Wildfire",
      "Natural disasters/weather",
      "Infrastructure",
      "Student safety",
      "Other"
    )
  )
)

glimpse(df)
```

## Closure data quality checks

```{r closure-balance-check}
df |>
  mutate(
    primary_cause_sum = rowSums(pick(all_of(primary_cause_vars)), na.rm = TRUE),
    total_minus_primary_causes = days_total - primary_cause_sum
  ) |>
  count(total_minus_primary_causes, sort = TRUE) |>
  slice_head(n = 20)
```

```{r student-safety-subcategory-check}
df |>
  mutate(
    student_safety_subtotal = coalesce(days_quarantine, 0) + coalesce(days_safety_other, 0),
    student_safety_difference = days_student_safety - student_safety_subtotal
  ) |>
  count(student_safety_difference, sort = TRUE) |>
  slice_head(n = 20)
```

```{r closure-school-year-duplicates}
df |>
  count(cds_code, closure_year, sort = TRUE) |>
  filter(n > 1) |>
  slice_head(n = 20)
```

## Step 1: Is wildfire a major driver of instructional disruption?

```{r statewide-yearly-summaries}
statewide_by_year <- df |>
  group_by(year) |>
  summarise(
    schools = n_distinct(cds_code),
    schools_with_any_closure = n_distinct(cds_code[days_total > 0]),
    total_enrollment = sum(total_enrollment, na.rm = TRUE),
    across(all_of(primary_cause_vars), ~ sum(.x, na.rm = TRUE)),
    days_total = sum(days_total, na.rm = TRUE),
    .groups = "drop"
  ) |>
  mutate(
    primary_cause_sum = rowSums(pick(all_of(primary_cause_vars)), na.rm = TRUE),
    wildfire_share_of_total_days = safe_divide(days_wildfire, days_total),
    wildfire_share_of_primary_causes = safe_divide(days_wildfire, primary_cause_sum),
    focus_year = year %in% analysis_years
  )

statewide_long <- statewide_by_year |>
  select(year, all_of(primary_cause_vars)) |>
  pivot_longer(
    cols = all_of(primary_cause_vars),
    names_to = "cause",
    values_to = "closure_days"
  ) |>
  left_join(cause_lookup, by = "cause")

statewide_by_year |>
  arrange(desc(days_wildfire)) |>
  select(
    year,
    schools,
    days_wildfire,
    days_total,
    wildfire_share_of_total_days,
    wildfire_share_of_primary_causes
  ) |>
  slice_head(n = 10)
```

```{r wildfire-vs-other-causes-line}
ggplot(
  statewide_long,
  aes(x = year, y = closure_days, color = cause_label, group = cause_label)
) +
  geom_line(linewidth = 1) +
  geom_point(size = 1.8) +
  geom_vline(xintercept = analysis_years, linetype = "dashed", color = "grey55", linewidth = 0.4) +
  scale_color_manual(values = cause_palette, name = NULL) +
  scale_x_continuous(breaks = year_breaks(statewide_long$year)) +
  scale_y_continuous(labels = label_comma()) +
  labs(
    title = "Statewide school closure days by cause",
    subtitle = "Dashed lines mark 2017-18 and 2018-19, the absenteeism comparison years",
    x = "Academic-year ending year",
    y = "Statewide closure days"
  ) +
  theme(
    legend.position = "bottom",
    panel.grid.minor = element_blank()
  )
```

```{r wildfire-share-stacked-bar}
statewide_share <- statewide_long |>
  group_by(year) |>
  mutate(
    year_primary_total = sum(closure_days, na.rm = TRUE),
    share = safe_divide(closure_days, year_primary_total)
  ) |>
  ungroup() |>
  filter(!is.na(share))

ggplot(
  statewide_share,
  aes(x = factor(year), y = share, fill = cause_label)
) +
  geom_col(width = 0.86, color = "white", linewidth = 0.15) +
  scale_fill_manual(values = cause_palette, name = NULL) +
  scale_y_continuous(labels = label_percent()) +
  labs(
    title = "Share of statewide school closure days by cause",
    subtitle = "Each bar sums to 100% across the primary documented closure causes",
    x = "Academic-year ending year",
    y = "Share of primary closure days"
  ) +
  theme(
    legend.position = "bottom",
    panel.grid.minor = element_blank(),
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
```

```{r total-disruption-wildfire-highlight}
total_with_wildfire_component <- statewide_by_year |>
  transmute(
    year,
    Wildfire = days_wildfire,
    `Non-wildfire days` = pmax(days_total - days_wildfire, 0)
  ) |>
  pivot_longer(
    cols = -year,
    names_to = "component",
    values_to = "closure_days"
  ) |>
  mutate(
    component = factor(component, levels = c("Non-wildfire days", "Wildfire"))
  )

ggplot(
  total_with_wildfire_component,
  aes(x = factor(year), y = closure_days, fill = component)
) +
  geom_col(width = 0.86) +
  scale_fill_manual(
    values = c(
      "Non-wildfire days" = "#D9D6CF",
      "Wildfire" = "#C7432B"
    ),
    name = NULL
  ) +
  scale_y_continuous(labels = label_comma()) +
  labs(
    title = "Total statewide closure days, with wildfire highlighted",
    subtitle = "Bar height uses days_total; red segment is days_wildfire",
    x = "Academic-year ending year",
    y = "Statewide closure days"
  ) +
  theme(
    legend.position = "bottom",
    panel.grid.minor = element_blank(),
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
```

```{r wildfire-reach-by-year}
wildfire_reach_by_year <- df |>
  group_by(year) |>
  summarise(
    schools = n_distinct(cds_code),
    schools_with_wildfire_closure = n_distinct(cds_code[days_wildfire > 0]),
    pct_schools_with_wildfire_closure = safe_divide(schools_with_wildfire_closure, schools),
    wildfire_days = sum(days_wildfire, na.rm = TRUE),
    .groups = "drop"
  )

ggplot(
  wildfire_reach_by_year,
  aes(x = year, y = pct_schools_with_wildfire_closure)
) +
  geom_area(fill = "#C7432B", alpha = 0.22) +
  geom_line(color = "#C7432B", linewidth = 1.1) +
  geom_point(color = "#C7432B", size = 1.8) +
  geom_vline(xintercept = analysis_years, linetype = "dashed", color = "grey55", linewidth = 0.4) +
  scale_x_continuous(breaks = year_breaks(wildfire_reach_by_year$year)) +
  scale_y_continuous(labels = label_percent(), limits = c(0, NA)) +
  labs(
    title = "Percent of schools with wildfire closures",
    x = "Academic-year ending year",
    y = "Percent of schools"
  ) +
  theme(panel.grid.minor = element_blank())
```

## Step 2: Is the burden uneven across counties, districts, and schools?

This section focuses on 2017-18 and 2018-19 to match the absenteeism files.

```{r closure-window}
closure_window <- df |>
  filter(year %in% analysis_years)

county_window_burden <- closure_window |>
  group_by(county) |>
  summarise(
    school_year_rows = n(),
    schools = n_distinct(cds_code),
    districts = n_distinct(district_name),
    total_enrollment_school_years = sum(total_enrollment, na.rm = TRUE),
    wildfire_days = sum(days_wildfire, na.rm = TRUE),
    total_closure_days = sum(days_total, na.rm = TRUE),
    wildfire_student_days = sum(wildfire_student_days, na.rm = TRUE),
    avg_wildfire_days_per_school_year = safe_divide(wildfire_days, school_year_rows),
    wildfire_share_of_total = safe_divide(wildfire_days, total_closure_days),
    .groups = "drop"
  ) |>
  filter(!is.na(county), county != "")

district_window_burden <- closure_window |>
  group_by(county, district_name) |>
  summarise(
    school_year_rows = n(),
    schools = n_distinct(cds_code),
    total_enrollment_school_years = sum(total_enrollment, na.rm = TRUE),
    wildfire_days = sum(days_wildfire, na.rm = TRUE),
    total_closure_days = sum(days_total, na.rm = TRUE),
    wildfire_student_days = sum(wildfire_student_days, na.rm = TRUE),
    avg_wildfire_days_per_school_year = safe_divide(wildfire_days, school_year_rows),
    wildfire_share_of_total = safe_divide(wildfire_days, total_closure_days),
    .groups = "drop"
  ) |>
  filter(!is.na(district_name), district_name != "")

school_window_burden <- closure_window |>
  group_by(cds_code, county, district_name, school_site) |>
  summarise(
    school_year_rows = n(),
    years_observed = n_distinct(year),
    total_enrollment_school_years = sum(total_enrollment, na.rm = TRUE),
    wildfire_days = sum(days_wildfire, na.rm = TRUE),
    total_closure_days = sum(days_total, na.rm = TRUE),
    wildfire_student_days = sum(wildfire_student_days, na.rm = TRUE),
    avg_wildfire_days_per_school_year = safe_divide(wildfire_days, school_year_rows),
    wildfire_share_of_total = safe_divide(wildfire_days, total_closure_days),
    .groups = "drop"
  ) |>
  filter(!is.na(cds_code))

list(
  counties_with_wildfire = county_window_burden |> filter(wildfire_days > 0) |> nrow(),
  districts_with_wildfire = district_window_burden |> filter(wildfire_days > 0) |> nrow(),
  schools_with_wildfire = school_window_burden |> filter(wildfire_days > 0) |> nrow()
)
```

### County burden rankings

```{r county-window-rankings}
top_n_counties <- 20

county_window_burden |>
  arrange(desc(wildfire_student_days)) |>
  select(
    county,
    schools,
    districts,
    wildfire_days,
    wildfire_student_days,
    avg_wildfire_days_per_school_year,
    wildfire_share_of_total
  ) |>
  slice_head(n = top_n_counties)

county_window_burden |>
  slice_max(wildfire_student_days, n = top_n_counties, with_ties = FALSE) |>
  ggplot(aes(x = wildfire_student_days, y = fct_reorder(county, wildfire_student_days))) +
  geom_col(fill = "#C7432B", alpha = 0.9) +
  scale_x_continuous(labels = label_comma()) +
  labs(
    title = "Counties with the largest wildfire missed student-days",
    subtitle = "2017-18 and 2018-19 only",
    x = "Wildfire missed student-days",
    y = NULL
  ) +
  theme(panel.grid.minor = element_blank())
```

### District burden rankings

```{r district-window-rankings}
top_n_districts <- 25

district_window_burden |>
  arrange(desc(wildfire_student_days)) |>
  select(
    county,
    district_name,
    schools,
    wildfire_days,
    wildfire_student_days,
    avg_wildfire_days_per_school_year,
    wildfire_share_of_total
  ) |>
  slice_head(n = top_n_districts)

district_window_burden |>
  mutate(district_label = str_c(str_trunc(district_name, 42), " (", county, ")")) |>
  slice_max(wildfire_student_days, n = 18, with_ties = FALSE) |>
  ggplot(aes(x = wildfire_student_days, y = fct_reorder(district_label, wildfire_student_days))) +
  geom_col(fill = "#B95A3C", alpha = 0.9) +
  scale_x_continuous(labels = label_comma()) +
  labs(
    title = "Districts with the largest wildfire missed student-days",
    subtitle = "2017-18 and 2018-19 only",
    x = "Wildfire missed student-days",
    y = NULL
  ) +
  theme(panel.grid.minor = element_blank())
```

### School burden rankings

```{r school-window-rankings}
top_n_schools <- 25

top_wildfire_schools <- school_window_burden |>
  arrange(desc(wildfire_student_days)) |>
  select(
    county,
    district_name,
    school_site,
    years_observed,
    total_enrollment_school_years,
    wildfire_days,
    wildfire_student_days,
    avg_wildfire_days_per_school_year,
    wildfire_share_of_total
  ) |>
  slice_head(n = top_n_schools)

top_wildfire_schools

school_window_burden |>
  mutate(school_label = str_c(str_trunc(school_site, 42), " (", county, ")")) |>
  slice_max(wildfire_student_days, n = 18, with_ties = FALSE) |>
  ggplot(aes(x = wildfire_student_days, y = fct_reorder(school_label, wildfire_student_days))) +
  geom_col(fill = "#C7432B", alpha = 0.9) +
  scale_x_continuous(labels = label_comma()) +
  labs(
    title = "Schools with the largest wildfire missed student-days",
    subtitle = "2017-18 and 2018-19 only",
    x = "Wildfire missed student-days",
    y = NULL
  ) +
  theme(panel.grid.minor = element_blank())
```

### Concentration of wildfire burden

```{r wildfire-burden-concentration}
make_concentration <- function(data, unit_label) {
  data |>
    filter(wildfire_student_days > 0) |>
    arrange(desc(wildfire_student_days)) |>
    mutate(
      rank = row_number(),
      rank_share = rank / n(),
      cumulative_share = cumsum(wildfire_student_days) / sum(wildfire_student_days, na.rm = TRUE),
      unit_type = unit_label
    )
}

concentration_data <- bind_rows(
  make_concentration(school_window_burden, "School"),
  make_concentration(district_window_burden, "District"),
  make_concentration(county_window_burden, "County")
)

ggplot(
  concentration_data,
  aes(x = rank_share, y = cumulative_share, color = unit_type)
) +
  geom_line(linewidth = 1.1) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "grey65") +
  scale_x_continuous(labels = label_percent()) +
  scale_y_continuous(labels = label_percent()) +
  scale_color_manual(values = c("School" = "#C7432B", "District" = "#3F7D8C", "County" = "#5E6E46"), name = NULL) +
  labs(
    title = "Wildfire missed student-days are concentrated",
    subtitle = "Curves above the dashed equality line indicate uneven burden",
    x = "Share of affected units, ranked from highest to lowest burden",
    y = "Cumulative share of wildfire missed student-days"
  ) +
  theme(
    legend.position = "bottom",
    panel.grid.minor = element_blank()
  )
```

### County maps

```{r county-map-data}
load_ca_counties <- function() {
  if (!requireNamespace("sf", quietly = TRUE)) {
    message("Package 'sf' is not installed; skipping map chunks.")
    return(NULL)
  }

  county_zip_url <- "https://www2.census.gov/geo/tiger/GENZ2018/shp/cb_2018_us_county_500k.zip"
  zip_path <- tempfile(fileext = ".zip")
  unzip_dir <- tempfile()
  dir.create(unzip_dir)

  tryCatch(
    {
      download.file(county_zip_url, zip_path, mode = "wb", quiet = TRUE)
      unzip(zip_path, exdir = unzip_dir)
      shp_path <- list.files(unzip_dir, pattern = "\\.shp$", full.names = TRUE)[1]

      sf::st_read(shp_path, quiet = TRUE) |>
        filter(STATEFP == "06") |>
        sf::st_transform(3310)
    },
    error = function(e) {
      message("County boundary download/read failed; skipping maps. Error: ", conditionMessage(e))
      NULL
    }
  )
}

ca_counties <- load_ca_counties()

county_map_data <- if (!is.null(ca_counties)) {
  ca_counties |>
    left_join(county_window_burden, by = c("NAME" = "county"))
} else {
  NULL
}

if (is.null(county_map_data)) {
  tibble(map_status = "County map data unavailable; non-map EDA sections still render.")
} else {
  county_map_data |>
    sf::st_drop_geometry() |>
    summarise(
      counties = n(),
      counties_matched_to_closure_data = sum(!is.na(wildfire_student_days)),
      counties_with_wildfire = sum(coalesce(wildfire_student_days, 0) > 0)
    )
}
```

```{r wildfire-county-maps}
if (!is.null(county_map_data)) {
  p1 <- ggplot(county_map_data) +
    geom_sf(aes(fill = wildfire_student_days), color = "white", linewidth = 0.15) +
    scale_fill_gradient(
      low = "#F7E7D8",
      high = "#9E2F1D",
      labels = label_comma(),
      na.value = "#F1F1F1",
      name = "Missed\nstudent-days"
    ) +
    labs(
      title = "Wildfire missed student-days by county",
      subtitle = "2017-18 and 2018-19"
    ) +
    theme_void() +
    theme(legend.position = "right")

  p2 <- ggplot(county_map_data) +
    geom_sf(aes(fill = avg_wildfire_days_per_school_year), color = "white", linewidth = 0.15) +
    scale_fill_gradient(
      low = "#F7E7D8",
      high = "#9E2F1D",
      labels = label_number(accuracy = 0.01),
      na.value = "#F1F1F1",
      name = "Avg days\nper school-year"
    ) +
    labs(
      title = "Average wildfire closure days per school-year by county",
      subtitle = "2017-18 and 2018-19"
    ) +
    theme_void() +
    theme(legend.position = "right")

  print(p1)
  print(p2)
}
```

## Step 3: What does absenteeism look like in affected schools?

### Load CDE Absenteeism by Reason data

CDE file source: <https://www.cde.ca.gov/ds/ad/filesabr.asp>

```{r load-absenteeism-data}
absenteeism_col_names <- c(
  "academic_year",
  "aggregate_level",
  "county_code",
  "district_code",
  "school_code",
  "county_name",
  "district_name",
  "school_name",
  "charter",
  "dass",
  "reporting_category",
  "eligible_cumulative_enrollment",
  "count_students_one_or_more_absences",
  "average_days_absent",
  "total_days_absent",
  "excused_absences_percent",
  "unexcused_absences_percent",
  "out_of_school_suspension_absences_percent",
  "incomplete_independent_study_absences_percent",
  "excused_absences_count",
  "unexcused_absences_count",
  "out_of_school_suspension_absences_count",
  "incomplete_independent_study_absences_count"
)

absenteeism_numeric_vars <- c(
  "eligible_cumulative_enrollment",
  "count_students_one_or_more_absences",
  "average_days_absent",
  "total_days_absent",
  "excused_absences_percent",
  "unexcused_absences_percent",
  "out_of_school_suspension_absences_percent",
  "incomplete_independent_study_absences_percent",
  "excused_absences_count",
  "unexcused_absences_count",
  "out_of_school_suspension_absences_count",
  "incomplete_independent_study_absences_count"
)

absence_reason_count_vars <- c(
  "excused_absences_count",
  "unexcused_absences_count",
  "out_of_school_suspension_absences_count",
  "incomplete_independent_study_absences_count"
)

read_absenteeism_file <- function(url, closure_year) {
  readr::read_tsv(
    url,
    skip = 1,
    col_names = absenteeism_col_names,
    col_types = readr::cols(.default = readr::col_character()),
    na = c("", "NA")
  ) |>
    clean_names() |>
    mutate(
      closure_year = closure_year,
      source_url = url
    )
}

absenteeism_sources <- tibble(
  academic_year = analysis_year_labels,
  closure_year = analysis_years,
  url = c(
    "https://www3.cde.ca.gov/demo-downloads/attendance/absenteeismreason18.txt",
    "https://www3.cde.ca.gov/demo-downloads/attendance/absenteeismreason19.txt"
  )
)

absenteeism_raw <- absenteeism_sources |>
  mutate(data = map2(url, closure_year, read_absenteeism_file)) |>
  select(data) |>
  unnest(cols = data)

absenteeism_schools <- absenteeism_raw |>
  filter(
    aggregate_level == "S",
    reporting_category == "TA"
  ) |>
  mutate(
    suppressed_numeric_fields = rowSums(across(all_of(absenteeism_numeric_vars), ~ .x == "*"), na.rm = TRUE),
    county_code_clean = pad_code(county_code, 2),
    district_code_clean = pad_code(district_code, 5),
    school_code_clean = pad_code(school_code, 7),
    cds_code = if_else(
      !is.na(county_code_clean) & !is.na(district_code_clean) & !is.na(school_code_clean),
      str_c(county_code_clean, district_code_clean, school_code_clean),
      NA_character_
    ),
    across(all_of(absenteeism_numeric_vars), to_number),
    county_name = str_squish(str_to_title(county_name)),
    district_name = str_squish(district_name),
    school_name = str_squish(school_name),
    absence_reason_count_sum = rowSums(pick(all_of(absence_reason_count_vars)), na.rm = TRUE),
    absence_reason_count_difference = total_days_absent - absence_reason_count_sum,
    share_students_with_absence = safe_divide(
      count_students_one_or_more_absences,
      eligible_cumulative_enrollment
    ),
    avg_days_absent_among_absent = average_days_absent,
    avg_days_absent_per_eligible = safe_divide(
      total_days_absent,
      eligible_cumulative_enrollment
    ),
    excused_absence_share = safe_divide(excused_absences_count, total_days_absent),
    unexcused_absence_share = safe_divide(unexcused_absences_count, total_days_absent),
    suspension_absence_share = safe_divide(out_of_school_suspension_absences_count, total_days_absent),
    incomplete_independent_study_absence_share = safe_divide(
      incomplete_independent_study_absences_count,
      total_days_absent
    )
  )

glimpse(absenteeism_schools)
```

### Absenteeism data quality checks

```{r absenteeism-quality-checks}
absenteeism_schools |>
  summarise(
    rows = n(),
    school_years = n_distinct(str_c(cds_code, closure_year)),
    schools = n_distinct(cds_code),
    rows_with_suppressed_values = sum(suppressed_numeric_fields > 0, na.rm = TRUE),
    rows_missing_cds_code = sum(is.na(cds_code))
  )

absenteeism_schools |>
  count(cds_code, closure_year, sort = TRUE) |>
  filter(n > 1) |>
  slice_head(n = 20)

absenteeism_schools |>
  summarise(
    across(
      all_of(absenteeism_numeric_vars),
      ~ mean(is.na(.x)),
      .names = "missing_share_{.col}"
    )
  ) |>
  pivot_longer(everything(), names_to = "field", values_to = "missing_share") |>
  arrange(desc(missing_share)) |>
  slice_head(n = 20)

absenteeism_schools |>
  mutate(abs_reason_abs_diff = abs(absence_reason_count_difference)) |>
  summarise(
    rows = n(),
    median_abs_difference = median(abs_reason_abs_diff, na.rm = TRUE),
    p95_abs_difference = quantile(abs_reason_abs_diff, 0.95, na.rm = TRUE),
    max_abs_difference = max(abs_reason_abs_diff, na.rm = TRUE)
  )
```

### Join closure school-years to absenteeism school-years

```{r closure-absenteeism-join}
closure_school_year <- closure_window |>
  group_by(cds_code, closure_year) |>
  summarise(
    academic_year = first(academic_year),
    county = first(county),
    district_name = first(district_name),
    school_site = first(school_site),
    total_enrollment = max(total_enrollment, na.rm = TRUE),
    days_wildfire = sum(days_wildfire, na.rm = TRUE),
    days_total = sum(days_total, na.rm = TRUE),
    wildfire_student_days = sum(wildfire_student_days, na.rm = TRUE),
    wildfire_share_of_total = safe_divide(days_wildfire, days_total),
    any_wildfire = any(days_wildfire > 0, na.rm = TRUE),
    .groups = "drop"
  ) |>
  mutate(total_enrollment = if_else(is.infinite(total_enrollment), NA_real_, total_enrollment))

closure_absenteeism_joined <- closure_school_year |>
  left_join(
    absenteeism_schools |>
      select(
        cds_code,
        closure_year,
        absenteeism_academic_year = academic_year,
        county_name_absenteeism = county_name,
        district_name_absenteeism = district_name,
        school_name_absenteeism = school_name,
        charter,
        dass,
        eligible_cumulative_enrollment,
        count_students_one_or_more_absences,
        average_days_absent,
        total_days_absent,
        all_of(absence_reason_count_vars),
        share_students_with_absence,
        avg_days_absent_among_absent,
        avg_days_absent_per_eligible,
        excused_absence_share,
        unexcused_absence_share,
        suspension_absence_share,
        incomplete_independent_study_absence_share,
        suppressed_numeric_fields
      ),
    by = c("cds_code", "closure_year")
  )

affected_cutoff <- closure_absenteeism_joined |>
  filter(wildfire_student_days > 0, !is.na(avg_days_absent_per_eligible)) |>
  summarise(cutoff = quantile(wildfire_student_days, 0.75, na.rm = TRUE)) |>
  pull(cutoff)

if (length(affected_cutoff) == 0 || is.na(affected_cutoff)) {
  affected_cutoff <- Inf
}

closure_absenteeism_joined <- closure_absenteeism_joined |>
  mutate(
    burden_group = case_when(
      wildfire_student_days <= 0 ~ "No wildfire",
      wildfire_student_days >= affected_cutoff ~ "Affected: top 25%",
      TRUE ~ "Affected: lower 75%"
    ),
    burden_group = factor(
      burden_group,
      levels = c("No wildfire", "Affected: lower 75%", "Affected: top 25%")
    )
  )

join_coverage <- tibble(
  metric = c(
    "closure_school_year_rows",
    "closure_school_year_rows_matched_to_absenteeism",
    "closure_school_year_match_rate",
    "unmatched_closure_school_year_rows",
    "absenteeism_school_year_rows_not_in_closure_data"
  ),
  value = c(
    nrow(closure_school_year),
    sum(!is.na(closure_absenteeism_joined$eligible_cumulative_enrollment)),
    safe_divide(
      sum(!is.na(closure_absenteeism_joined$eligible_cumulative_enrollment)),
      nrow(closure_school_year)
    ),
    nrow(anti_join(closure_school_year, absenteeism_schools, by = c("cds_code", "closure_year"))),
    nrow(anti_join(absenteeism_schools, closure_school_year, by = c("cds_code", "closure_year")))
  )
)

join_coverage
```

```{r unmatched-school-year-examples}
anti_join(closure_school_year, absenteeism_schools, by = c("cds_code", "closure_year")) |>
  select(cds_code, closure_year, county, district_name, school_site, days_wildfire, days_total) |>
  arrange(desc(days_wildfire), county, district_name, school_site) |>
  slice_head(n = 20)

anti_join(absenteeism_schools, closure_school_year, by = c("cds_code", "closure_year")) |>
  select(cds_code, closure_year, county_name, district_name, school_name, eligible_cumulative_enrollment) |>
  arrange(county_name, district_name, school_name) |>
  slice_head(n = 20)
```

### Absenteeism by wildfire burden group

```{r absenteeism-by-burden-group-summary}
burden_group_summary <- closure_absenteeism_joined |>
  filter(!is.na(avg_days_absent_per_eligible)) |>
  group_by(burden_group) |>
  summarise(
    school_year_rows = n(),
    schools = n_distinct(cds_code),
    wildfire_days = sum(days_wildfire, na.rm = TRUE),
    wildfire_student_days = sum(wildfire_student_days, na.rm = TRUE),
    eligible_cumulative_enrollment = sum(eligible_cumulative_enrollment, na.rm = TRUE),
    count_students_one_or_more_absences = sum(count_students_one_or_more_absences, na.rm = TRUE),
    total_days_absent = sum(total_days_absent, na.rm = TRUE),
    share_students_with_absence = safe_divide(
      count_students_one_or_more_absences,
      eligible_cumulative_enrollment
    ),
    avg_days_absent_per_eligible = safe_divide(total_days_absent, eligible_cumulative_enrollment),
    avg_days_absent_among_absent = safe_divide(
      total_days_absent,
      count_students_one_or_more_absences
    ),
    .groups = "drop"
  )

burden_group_summary
```

```{r absenteeism-by-burden-group-boxplots}
closure_absenteeism_joined |>
  filter(!is.na(avg_days_absent_per_eligible)) |>
  ggplot(aes(x = burden_group, y = avg_days_absent_per_eligible, fill = burden_group)) +
  geom_boxplot(outlier.alpha = 0.15) +
  scale_fill_manual(values = burden_palette, guide = "none") +
  scale_y_continuous(labels = label_number(accuracy = 0.1)) +
  labs(
    title = "Average days absent per eligible student by wildfire burden group",
    subtitle = "School-year rows, 2017-18 and 2018-19",
    x = NULL,
    y = "Days absent per eligible student"
  ) +
  theme(
    panel.grid.minor = element_blank(),
    axis.text.x = element_text(angle = 15, hjust = 1)
  )

closure_absenteeism_joined |>
  filter(!is.na(share_students_with_absence)) |>
  ggplot(aes(x = burden_group, y = share_students_with_absence, fill = burden_group)) +
  geom_boxplot(outlier.alpha = 0.15) +
  scale_fill_manual(values = burden_palette, guide = "none") +
  scale_y_continuous(labels = label_percent()) +
  labs(
    title = "Share of eligible students with one or more absences",
    subtitle = "School-year rows, 2017-18 and 2018-19",
    x = NULL,
    y = "Share with one or more absences"
  ) +
  theme(
    panel.grid.minor = element_blank(),
    axis.text.x = element_text(angle = 15, hjust = 1)
  )
```

```{r absence-reason-composition-by-burden}
reason_labels <- c(
  "excused_absences_count" = "Excused",
  "unexcused_absences_count" = "Unexcused",
  "out_of_school_suspension_absences_count" = "Suspension",
  "incomplete_independent_study_absences_count" = "Incomplete independent study"
)

reason_composition_by_group <- closure_absenteeism_joined |>
  filter(!is.na(total_days_absent), total_days_absent > 0) |>
  group_by(burden_group) |>
  summarise(
    total_days_absent = sum(total_days_absent, na.rm = TRUE),
    across(all_of(absence_reason_count_vars), ~ sum(.x, na.rm = TRUE)),
    .groups = "drop"
  ) |>
  pivot_longer(
    cols = all_of(absence_reason_count_vars),
    names_to = "reason",
    values_to = "absence_days"
  ) |>
  mutate(
    reason = recode(reason, !!!reason_labels),
    reason_share = safe_divide(absence_days, total_days_absent)
  )

ggplot(
  reason_composition_by_group,
  aes(x = burden_group, y = reason_share, fill = reason)
) +
  geom_col(width = 0.78, color = "white", linewidth = 0.2) +
  scale_y_continuous(labels = label_percent()) +
  scale_fill_manual(
    values = c(
      "Excused" = "#477998",
      "Unexcused" = "#D0A03A",
      "Suspension" = "#7C4D3A",
      "Incomplete independent study" = "#7D7A78"
    ),
    name = NULL
  ) +
  labs(
    title = "Absence reason composition by wildfire burden group",
    subtitle = "Reason counts are summed within group before calculating shares",
    x = NULL,
    y = "Share of total absence days"
  ) +
  theme(
    legend.position = "bottom",
    panel.grid.minor = element_blank(),
    axis.text.x = element_text(angle = 15, hjust = 1)
  )
```

```{r wildfire-burden-vs-absenteeism-scatter}
closure_absenteeism_joined |>
  filter(wildfire_student_days > 0, !is.na(avg_days_absent_per_eligible)) |>
  ggplot(aes(x = wildfire_student_days, y = avg_days_absent_per_eligible)) +
  geom_point(aes(size = eligible_cumulative_enrollment), alpha = 0.35, color = "#C7432B") +
  geom_smooth(method = "loess", se = FALSE, color = "#4A5568") +
  scale_x_continuous(labels = label_comma()) +
  scale_size_continuous(labels = label_comma(), guide = "none") +
  labs(
    title = "Wildfire burden and absenteeism among affected school-years",
    subtitle = "Descriptive only; this is not a causal estimate",
    x = "Wildfire missed student-days",
    y = "Average days absent per eligible student"
  ) +
  theme(panel.grid.minor = element_blank())
```

### High-burden school profiles

```{r high-burden-school-profiles}
school_absenteeism_profile <- closure_absenteeism_joined |>
  group_by(cds_code, county, district_name, school_site) |>
  summarise(
    years_observed = n_distinct(closure_year),
    wildfire_days = sum(days_wildfire, na.rm = TRUE),
    days_total = sum(days_total, na.rm = TRUE),
    wildfire_student_days = sum(wildfire_student_days, na.rm = TRUE),
    total_enrollment_school_years = sum(total_enrollment, na.rm = TRUE),
    eligible_cumulative_enrollment = sum(eligible_cumulative_enrollment, na.rm = TRUE),
    count_students_one_or_more_absences = sum(count_students_one_or_more_absences, na.rm = TRUE),
    total_days_absent = sum(total_days_absent, na.rm = TRUE),
    across(all_of(absence_reason_count_vars), ~ sum(.x, na.rm = TRUE)),
    .groups = "drop"
  ) |>
  mutate(
    wildfire_share_of_total = safe_divide(wildfire_days, days_total),
    avg_days_absent_per_eligible = safe_divide(total_days_absent, eligible_cumulative_enrollment),
    share_students_with_absence = safe_divide(
      count_students_one_or_more_absences,
      eligible_cumulative_enrollment
    ),
    avg_days_absent_among_absent = safe_divide(
      total_days_absent,
      count_students_one_or_more_absences
    ),
    excused_absence_share = safe_divide(excused_absences_count, total_days_absent),
    unexcused_absence_share = safe_divide(unexcused_absences_count, total_days_absent),
    suspension_absence_share = safe_divide(out_of_school_suspension_absences_count, total_days_absent),
    incomplete_independent_study_absence_share = safe_divide(
      incomplete_independent_study_absences_count,
      total_days_absent
    )
  )

top_school_absenteeism_profiles <- school_absenteeism_profile |>
  filter(wildfire_student_days > 0) |>
  arrange(desc(wildfire_student_days)) |>
  select(
    county,
    district_name,
    school_site,
    years_observed,
    wildfire_days,
    wildfire_student_days,
    eligible_cumulative_enrollment,
    avg_days_absent_per_eligible,
    share_students_with_absence,
    excused_absence_share,
    unexcused_absence_share,
    suspension_absence_share,
    incomplete_independent_study_absence_share
  ) |>
  slice_head(n = 25)

top_school_absenteeism_profiles
```

```{r high-burden-school-absenteeism-charts}
top_school_plot_data <- school_absenteeism_profile |>
  filter(wildfire_student_days > 0) |>
  slice_max(wildfire_student_days, n = 15, with_ties = FALSE) |>
  mutate(school_label = str_c(str_trunc(school_site, 38), " (", county, ")"))

ggplot(
  top_school_plot_data,
  aes(x = avg_days_absent_per_eligible, y = fct_reorder(school_label, wildfire_student_days))
) +
  geom_col(fill = "#3F7D8C", alpha = 0.9) +
  scale_x_continuous(labels = label_number(accuracy = 0.1)) +
  labs(
    title = "Absenteeism in the highest wildfire-burden schools",
    subtitle = "Top schools ranked by wildfire missed student-days, 2017-18 and 2018-19",
    x = "Average days absent per eligible student",
    y = NULL
  ) +
  theme(panel.grid.minor = element_blank())

top_school_reason_long <- top_school_plot_data |>
  select(school_label, total_days_absent, all_of(absence_reason_count_vars)) |>
  pivot_longer(
    cols = all_of(absence_reason_count_vars),
    names_to = "reason",
    values_to = "absence_days"
  ) |>
  mutate(
    reason = recode(reason, !!!reason_labels),
    reason_share = safe_divide(absence_days, total_days_absent)
  )

ggplot(
  top_school_reason_long,
  aes(x = reason_share, y = fct_reorder(school_label, reason_share, .fun = sum), fill = reason)
) +
  geom_col(width = 0.8, color = "white", linewidth = 0.15) +
  scale_x_continuous(labels = label_percent()) +
  scale_fill_manual(
    values = c(
      "Excused" = "#477998",
      "Unexcused" = "#D0A03A",
      "Suspension" = "#7C4D3A",
      "Incomplete independent study" = "#7D7A78"
    ),
    name = NULL
  ) +
  labs(
    title = "Absence reason composition in high-burden schools",
    x = "Share of total absence days",
    y = NULL
  ) +
  theme(
    legend.position = "bottom",
    panel.grid.minor = element_blank()
  )
```

## Optional: closure burden and test outcomes

This is descriptive only and not a causal model.

```{r optional-test-score-scatter}
test_score_vars <- c(
  "percentage_standard_met_and_above_3_math",
  "percentage_standard_met_and_above_3_english",
  "percentage_standard_met_and_above_8_math",
  "percentage_standard_met_and_above_8_english",
  "percentage_standard_met_and_above_11_math",
  "percentage_standard_met_and_above_11_english"
)

test_score_long <- df |>
  filter(year %in% analysis_years) |>
  select(cds_code, year, county, days_wildfire, wildfire_student_days, any_of(test_score_vars)) |>
  mutate(across(any_of(test_score_vars), to_number)) |>
  pivot_longer(
    cols = any_of(test_score_vars),
    names_to = "test_metric",
    values_to = "percent_met_or_above"
  ) |>
  filter(!is.na(percent_met_or_above))

ggplot(
  test_score_long,
  aes(x = wildfire_student_days, y = percent_met_or_above)
) +
  geom_jitter(width = 0.15, height = 0, alpha = 0.08, size = 0.7) +
  geom_smooth(method = "loess", se = FALSE, color = "#C7432B") +
  facet_wrap(vars(test_metric), scales = "free_y") +
  scale_x_continuous(labels = label_comma()) +
  labs(
    title = "Exploratory relationship between wildfire burden and test performance",
    subtitle = "Descriptive only; interpret cautiously because schools differ in many other ways",
    x = "Wildfire missed student-days",
    y = "Percent meeting or exceeding standard"
  ) +
  theme(panel.grid.minor = element_blank())
```
